\mikexercise{Define the  \emph{orbit count} of a deterministic pof automaton to be the number of orbits of reachable states. For a language, we can consider the set of  deterministic pof automata that recognise it, and have minimal orbit count for this property. Show that this set can contain automata that are non-isomorphic. Here, an isomorphism between two automata is an equivariant bijection 
\begin{align*}
\text{reachable states of $\Aa$}
\quad \stackrel f \longrightarrow \quad 
\text{reachable states of $\Bb$}
\end{align*}
such that for every input word, applying $f$ to the state of $\Aa$ after reading a word gives the state of $\Bb$ after reading the same word. 
} 
{
    Consider the language of one-letter words, over the alphabet $\atoms$. One automaton does not store any atoms, i.e.~it has three states ``initial'', ``one letter'', and ``at least two letters''. This corresponds to disjoint union of three copies of $\atoms^0$. 
    An alternative automaton stores the last letter in its state. Therefore, the two automata have state spaces: 
    \begin{align}
        \myunderbrace{\atoms^0 + \atoms^0 + \atoms^0}{first automaton}
        \quad \text{and} \quad
        \myunderbrace{\atoms^0 + \atoms^1 + \atoms^0}{second automaton}.
    \end{align}
    In both cases, the state space has three orbits. Clearly the first automaton is better, but this is not reflected in the orbit count alone. One can also show that an automaton with fewer orbits cannot do the trick. Indeed, the initial state must be its own orbit $\atoms^0$. This state must be rejecting, since the empty word is not in the language. The other orbit must be accepting (since otherwise the automaton would reject all words). In order to reject a word of length two or more, the automaton would need to return to the initial state, which would mean that it would accept some even longer words.
}

\mikexercise{
    \label{ex:hitting-set-deterministic-generalised}
    Let $d, e \in \set{1,2,\ldots}$. Show that the language 
    \begin{align*}
    \setbuild{w \in {\atoms \choose d}^*}{there is a hitting  set of at most $e$ atoms}
    \end{align*}
    is recognised by a deterministic spof automaton. 
}{}

\mikexercise{How many equivariant functions are there of type 
\begin{align*}
{\atoms \choose d} \to {\atoms \choose e}
\end{align*}
for $d, e \in \set{0,1,\ldots}$?
}{
One possibility is the identity function, which arises when $d=e$. We will show that this is the only possibility. 
There cannot be any such function when $e > d$, because otherwise the output would need to contain an atom that is not present in the input. There cannot be any such function There cannot be any such function when $e < d$, because this would require choosing some atom, and this cannot be done by the argument in Example~\ref{ex:choice-spof}. Finally, when $d=e$, the only possibility is the identity function, because otherwise the output would need to contain some atom that is not in the input.
}

\mikexercise{\label{ex:intersection-language} Find a deterministic spof automaton for the language
\begin{align*}
    \setbuild{ w \in {\atoms \choose 3}^*}{some atom $a$ appears in all letters}.
    \end{align*}}{The automaton computes the intersection of all letters, and accepts if this intersection is nonempty. The possible }

    \mikexercise{\label{ex:union-language} Find a deterministic spof automaton for the language
\begin{align*}
    \setbuild{ w \in {\atoms \choose 3}^*}{there are at most distinct 5 atoms used in the word}.
    \end{align*}}{The automaton computes the union of all letters, up to threshold 5. }

\mikexercise{\label{ex:two-atoms-from-first-everywhere} Find a deterministic spof automaton for the language
    \begin{align*}
        \setbuild{ w \in {\atoms \choose 3}^*}{some atom from the set in the first letter appears \\ an even number of times in the remaining letters}.
        \end{align*}
}{}

\mikexercise{\label{ex:some-two-atoms-everywhere} Find a deterministic spof automaton for the language:
\begin{align*}
\setbuild{ w \in {\atoms \choose 2}^*}{there exist $a,b \in \atoms$ such that every letter in $w$ intersects $\set{a,b}$}.
\end{align*}
}{
Consider the first two letters in the input string that are not equal to each other, which are sets $x$ and $y$ of size two. 
\begin{enumerate}
    \item If the sets are disjoint, then the only candidates for $a,b$ are from $x \cup y$. Then, we can use the same kind of solution as in Exercise~\ref{ex:two-atoms-from-first-everywhere}.
    \item Otherwise, the only candidates for $a$ and $b$ are: 
    \begin{enumerate}
        \item the two atoms that are in the symmetric difference $(x \setminus y) \cup (y \setminus x)$; or 
        \item the atom in the intersection, and some other atom. 
    \end{enumerate}
\end{enumerate}
}

\mikexercise{\label{ex:group-colcombet} Consider a spof group, i.e.~the underlying set is a spof, and the group operation is equivariant. Show that such a group must be  finite.} { This exercise is based on~\cite[Lemma 2.14]{DBLP:journals/corr/ColcombetLP14}.
Consider the least support of the multiplication operation in the group. This least support also supports the universe of the group, and the inverse operation $g \mapsto g^{-1}$. For an element $g$ of the group, define $[g]$ to be the set of atoms that are in the least support of $g$ but are not in the least support of the multiplication operation of the group. 
If a set of atoms supports $g,h $ and the multiplication operation, then it also supports the product $gh$. It follows that 
\begin{align}\label{eq:group-homo}
 [g h] \subseteq [g] \cup [h] .
\end{align}
For the same reasons, we have
\begin{align}
 \label{eq:group-homo2}
 [g^{-1}] = [g] .
\end{align}
Take some $g$ in the group which maximises the size $[g]$. Such a maximum exists, since the size of $[g]$ depends on that $\bar a$-orbit of $g$, of which there are finitely many. Since we are dealing with the equality atoms, we can choose an atom automorphism $\pi$ so that 
\begin{align}\label{eq:group-abc}
 \pi([g]) \cap [g] = \emptyset.
\end{align}
 We have
\begin{align*}
 g = \pi(g) \pi(g)^{-1} g.
\end{align*}
Combining this with~\eqref{eq:group-homo}, we get 
\begin{align*}
 [g] \subseteq [\pi(g)] \cup [\pi(g)^{-1} g]
\end{align*}
Combining this with~\eqref{eq:group-abc}, we get
\begin{align*}
 [g] \subseteq [\pi(g)^{-1}g]
\end{align*}
By maximality of $[g]$ the above is actually an equality, i.e.
\begin{align}
 \label{eq:group-homo3}
 [g] \subseteq [\pi(g)^{-1}g]
\end{align}
The same proof also yields
\begin{align}
 \label{eq:group-homo4}
 [\pi(g)] \subseteq [g^{-1}\pi(g)]
\end{align}
 Using a similar reasoning applied to
\begin{align*}
 \pi(g)^{-1} = g^{-1} \pi(g) \pi(g)^{-1}
\end{align*}
we conclude that 
\begin{align*}
 [\pi(g)] \stackrel {\text{\eqref{eq:group-homo4}}} \subseteq [ g^{-1} \pi(g)] \stackrel {\text{\eqref{eq:group-homo2}}}= [\pi(g)^{-1}g] \stackrel {\text{\eqref{eq:group-homo3}}}= [g].
\end{align*}
From~\eqref{eq:group-abc} it follows that $[\pi(g)]$ is empty. Therefore, $[g]$ must also be empty, since $[\_]$ commutes with $\bar a$-automorphisms. 
By maximality of $[g]$ it follows that all elements of the group have value $\emptyset$ under $[\_]$ which implies that all elements of the group are supported by $\bar a$. In an orbit-finite set there can only be finitely many elements with a given support (Exercise~\ref{ex:finitely-many-supported-by-one-tuple}). Therefore, the group is finite.
The same proof would work for some other atoms, e.g.~$\qatom$.}






\mikexercise{
 Let $X$ be a spof.  Show that if $f : X \to X$ is an equivariant surjective function, then $f$ is a bijection.
}
{
 
}


\mikexercise{Consider a chain 
\begin{align*}
X_0 \stackrel {f_1} \twoheadrightarrow X_1 \stackrel {f_2} \twoheadrightarrow X_2 \stackrel {f_3} \twoheadrightarrow \ldots
\stackrel {f_n} \twoheadrightarrow X_n
\end{align*}
of equivariant surjective function between spof sets.
Show that the length of this chain is bounded by a polynomial of the following two parameters of the first set $X_0$: the orbit count, and the atom dimension.
}{}

